---
title: "Quantifying collection bias related to accessibility"
author: "Alex Zizka"
date: "16 September 2017"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, 
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```

## Library setup
You will need the following R libraries for this exercise, just copy the code chunk into you R console to load them. You might need to install some of them separately.

```{r}
library(tidyverse)
library(sampbias)
```

# Solutions

## 1. Load the example distribution data from mammals on Borneo ("borneo_mammals.csv")
```{r, message = F, warning = F}
occ <- read_csv("inst/occurrence_records_clean.csv")%>%
  mutate(decimallongitude = decimalLongitude)%>%
  mutate(decimallatitude = decimalLatitude)
```

## 2. Run sampbias with the default settings. (SamplingBias)
```{r, message = F, warning = F}
bias.out <- SamplingBias(x = occ)
```

## 3. Look at the run summary and visualize the results. How informative are the results?
```{r, message = F, warning = F}
#summarize results
summary(bias.out)

#Visualize
plot(bias.out)
```

## 4. Explore `?SamplingBias` and try to change the relevant arguments to improve the results. Summarize and visualize again.
```{r, message = F, warning = F, eval = F}
bias.det <- SamplingBias(x = occ, res = 0.1)

#summarize results
summary(bias.det)

#Visualize
par(mfrow = c(3,2))
plot(bias.det)
```

## 5. Explore the Bio-Dem app to explore the effect of political factos on data collections in your group. 
http://bio-dem.surge.sh/

