---
title: "Species richness maps and range size estimation from point occurrences"
output: 
  html_document:
    theme: readable
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, 
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```

```{r}
library(readr)
library(ggplot2)
library(magrittr)
library(speciesgeocodeR)
library(raster)
library(rgdal)
```

## 1. Load occurrence data and visualize
```{r}
dat <- read_csv("inst/occurrence_records_clean.csv")

# Visualize
world.inp  <- map_data("world")

ggplot()+
  geom_map(data=world.inp, map=world.inp, aes(x=long, y=lat, map_id=region), fill = "grey80")+
  xlim(min(dat$decimalLongitude, na.rm = T), max(dat$decimalLongitude, na.rm = T))+
  ylim(min(dat$decimalLatitude, na.rm = T), max(dat$decimalLatitude, na.rm = T))+
  geom_point(data = dat, aes(x = decimalLongitude, y = decimalLatitude),
             colour = "darkblue", size = 1)+
  coord_fixed()+
  theme_bw()+
  theme(axis.title = element_blank())
```

## 2. Visualize the number of occurrence records and the number of species in a regular-sized grid (`RichnessGrid`)

For a quick overview you can first visualize pattern in a lat/long grid.

```{r}
# The number of occurrences per grid cell
gocc <- RichnessGrid(dat, type = "abu")

plot(gocc)

# Or more sofisticated plotting
# Prepare for plotting
plo_gocc <- data.frame(rasterToPoints(gocc))

# Visualize
ggplot()+
  geom_map(data = world.inp, map = world.inp, aes(x = long, y = lat, map_id = region), fill = "grey80")+
  xlim(min(dat$decimalLongitude, na.rm = T), max(dat$decimalLongitude, na.rm = T))+
  ylim(min(dat$decimalLatitude, na.rm = T), max(dat$decimalLatitude, na.rm = T))+
  geom_raster(data = plo_gocc, aes(x = x, y = y, fill = layer))+
  #scale_fill_viridis(name = "Species", direction = -1)+
  coord_fixed()+
  theme_bw()+
  ggtitle("Number of occurrences")+
  theme(axis.title = element_blank())

# The number of species per grid cell
gri <- RichnessGrid(dat)

plot(gri)
```


#3. Convex hull ranges per species

As the simplest model of species distribution we can use convex hull to fill the gaps.

```{r}
# Calculate ranges
ranges <- CalcRange(dat)

# Visualize
plot(ranges)

# Write to working directory
writeOGR(ranges, dsn = "inst", layer = "range_polygons", 
         driver = "ESRI Shapefile", overwrite_layer = TRUE)

```

## 4. Visualize species richness per grid cell based on the range polygons.

```{r}
# Create richness raster
r_ri <- RangeRichness(ranges, res = 0.5)

plot(r_ri)
```

## 5. Visualise species richenss in an equal area raster

So far we have displayed species richness in a lat/lon projected raster. This is OK to get a feeling for the data, but strictly speeaking wrong, since the actuala rea of a 1x1 degree cell varies with latitude. So, it is better to use an eaqual area pojection, such as the Lamberts projection.


```{r}
# Define projections
wgs1984 <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs")
behr <- CRS('+proj=cea +lon_0=0 +lat_ts=30 +x_0=0 +y_0=0 +datum=WGS84 +ellps=WGS84 +units=m +no_defs')

# reproject the occurrence records
pts <- dat[, c("decimalLongitude", "decimalLatitude")]%>%
  SpatialPoints(proj4string = wgs1984)%>%
  spTransform(behr)

# create a aqual area template in behrman projection
be <- raster(ncol = 360, nrow = 142, 
             xmn = -17367529, xmx = 17367529, 
             ymn = -6356742, ymx = 7348382,
             crs = behr)
be <- crop(be, extent(pts))

pts <- data.frame(dat$species, coordinates(pts))

# Equal area Occurrence number
eq_occgri <- RichnessGrid(x= pts, ras = be, type = "abu")
plot(eq_occgri)

# Equal area speces richness
eq_rigri <- RichnessGrid(x = pts, ras = be)
plot(eq_rigri)

# Range based species richness
ranges <- spTransform(ranges, behr)

eq_rri <- RangeRichness(x = ranges, ras = be)

plot(eq_rri)
```


6. Write to disk
```{r}
writeRaster(eq_rri, filename = "inst/equal_area_range_richness", 
            format = "ascii", overwrite = TRUE)
```

