---
title: "Range size and conservation assessment"
output: 
  html_document:
    theme: readable
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, 
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```
## Background
Species occurrence records can be used to approximate species ranges and generate preliminary conservation assessments. While comprehensive Red list assessments need a through case-by-case evaluation, preliminary assessments can help to speed up this process, by focussing on potentially threatened species.

## Objectives
After this exercise you will be abler to 
* Approximate species range sizes based on occurrence records
* Conduct an automated preliminary conservation assessment for multiple species based on occurrence records and Criterion B of the International Union for the conservation of Nature.

## Exercises
1. Calculate the area of the EOO for all species in your group (`CalcRangeSize`)
2. Do a preliminary conservation assessment of your group based on Criterion B using the ConR package. You can find a detailed tutorial [here](https://cran.r-project.org/web/packages/ConR/vignettes/my-vignette.html). (`IUCN.eval`)
3. Look at

## Possible questions for your project
* What is the maximum range size for a species in your group
* What does the distribution of range sizes look like? Is it normally distributed?
* What is the conservation status of your species following Criterion B?
* What does the regional assessment mean?
* What caveats could there be with the range based conservation assessment?

## Library setup
```{r}
library(speciesgeocodeR)
library(ConR)
library(tidyverse)
```

